---
title: "Gillespie Algorithm"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---
# Initialisation

Firstly, we need to install required packages.

```{r echo=FALSE}
require(GillespieSSA)
require(ggplot2)
require(dplyr)
require(foreach)
require(reshape2)
```

To do this you should use function `install.packages(package_name)`.

# Gene expression model. 

We will consider simple GEM defined as:

dm/dt = Sm - km * m
dn/dt = Sn * m - kn * n

To simulate model using Gillespie Algorithm we will use ssa function implemented in package GillespieSSA. Let's read the ssa's documentation `?ssa`.

> To simulate model we need to define:

* nu --- state change matrix (stoichometric matrix)
    * rows corresponds to variables 
    * columns corresponds to reactions
    * each element e_{i,j} of matrix tells what is an effect of reaction j on variable i 
* a --- propensity vector - prawdopodobienstwo przejscia pomiedzy stanami reakcji

```{r}
nu <- 
  matrix(
    c(1,  -1, 0, 0, 
      0, 0,  1, -1),
  nrow = 2,
  byrow = TRUE)
a <- 
  c("Sm", 
    "km*m",
    "Sn*m",
    "kn*n")
```

> Moreover we need to initiate parameters and variables state:

* x0 --- initial variables state 
* param --- vector of parameters values
* tf --- time of simulations


```{r}
x0 <- c(m = 1, n = 0)
parms <- c(Sm = 0.05, km = 0.005, Sn = 0.05, kn = 0.001)
tf <- 5000
```

Model simulation:
```{r}
method <- "D"
simName <- "Gene Expression Model"
out <- ssa(x0 = x0,
           a = a,
           nu = nu,
           parms = parms,
           tf = tf,
           method = method,
           simName = simName,
           verbose = FALSE,
           consoleInterval = 1)
```


```{r}
data <- data.frame(out$data)
colnames(data)[1] <- "t"
head(data)
```
Plotting mRNA
```{r}
g <- 
  ggplot(
    data = data,
    mapping = aes(x = t, y = m)) +
  geom_line() +
  xlab("mRNA molecules number") +
  coord_cartesian(ylim = c(0, 1.2*max(data$m)))
print(g)
```
Plotting mRNA and protein in time
```{r}
data <- data.frame(out$data)
colnames(data)[1] <- "t"
data.long <- 
  data %>%
  reshape2::melt(id = "t")
g <- 
  ggplot(
    data = data.long,
    mapping = aes(x = t, y = value, group = variable, color = variable)) +
  geom_line() +
  xlab("Molecules number") +
  coord_cartesian(ylim = c(0, 1.2*max(data$m, data$n)))
print(g)
```

# Task 1  
 Consider GEM with different initiate state values.
   
   * Initiate  in steady state
   * Initiate state m = 1, n = 0
   * Initiate state m = 10, n = 0
 
 Prepare code that simulates model multiple times and plot trajectories of molecules numbers.  
  
  * Calculate and plot mean and standard deviation of molecules in time. For mRNA compare results in steady state  with analytical solution of master equations.
  *  Calculate time of convergence to the steady state.
  *  Plot mRNA vs protein
 
 
# Task 2
> Lottka-Volter Model

 * R1: A + X --> 2X  
 * R2: X + Y --> 2Y
 * R3: Y --> *
 
 X -- rabbits 
 Y -- foxes 
 
 'reaction' rates: k1, k2, k3
 
 A -  paramter
 
 * Calulcate the propensity function and  stechiometric matrix
 * Caluclate steady state 
 * Simulate model multiple times starting from different states. What is a time of convergance ?
 * Prepare plot presenting population of rabbits vs foxes

# Task 3 
Zrób symulację następującego układu:

A + B → C

D → A + B

→ E

E + C → D

A → *

B → *

C → *

D → *

E → *

i znajdź taki rozkład prawdopodobieństw poszczególnych reakcji, który skutkuje dwoma scenariuszami:

układ znajduje się mniej więcej w stanie stacjonarnym
układ rośnie w nieskończoność
Pokaż wykresy stężeń dla wszystkich związków.

Jako stan początkowy przyjmij losowe ilości cząsteczek wszystkich związków z przedziału <2,10>.